Seasonality of submesoscale dynamics in the Kuroshio Extension

### 2016-07-11 writeup/rocha_etal.tex
Fixes some minor stuff

\begin{abstract}
    Two new high-resolution numerical simulations with embedded tides show a
    strong modulation in the governing near-surface dynamics at submesoscales
    (roughly 10-100 km) in the Kuroshio Extension. Consistent with recent studies, deep late-winter mixed layers
    are prone to baroclinic instabilities, and submesoscale turbulence
    prevails in late winter/early spring. While summertime
    re-stratification weakens submesoscale turbulence, it also enhances submesoscale inertia-gravity
     waves near the surface. In the Kuroshio Extension,
    inertia-gravity waves strongly dominate the submesoscale surface kinetic energy and
    sea-surface height variance in late summer/early fall.
\end{abstract}

@@

Recent interest in upper-ocean dynamics has focused on the strong seasonal
cycle of shallow baroclinic instabilities and their role in submesoscale (roughly 1-100 km)
turbulence and mesoscale modulation \citep{sasaki_etal2014,qiu_etal2014,
callies_etal2015, thompson_etal2016,buckingham_etal2016}. Contemporary studies
have also shown that inertia-gravity waves contribute significantly
to the near-surface variability at submesoscales \citep{richman_etal2012,
buhler_etal2014,rocha_etal2016}, but their seasonality has not been investigated.

@@

Inertia-gravity waves, on the other hand, peak in late summer/early fall,
when the upper ocean is strongly stratified. Thus there exists a
strong seasonal modulation in the dominant upper-ocean submesoscale dynamics:
submesoscale turbulence dominates the upper-ocean dynamics in late winter/early,
whereas submesoscale inertia-gravity waves prevail in late summer/early fall.
Because submesoscale turbulence is weakest in late summer/early fall,
the present results indicate that inertia-gravity waves account for most of the
summertime submesoscale SSH variability.

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
realistic numerical simulations. The outputs
analyzed here, LLC2160 (1/24$^\circ$) and LLC4320 (1/48$^\circ$),   are
forward Massachusetts Institute of Technology general circulation model (MITgcm)
numerical solutions on a LLC grid  \citep{forget_etal2015} with
90-vertical-levels. The coarser-resolution LLC simulation was
spun up from an ECCO2 adjoint-method state estimate, constrained to millions
of observations from 2009 through 2011. Both simulations were forced by
tides and 6-hourly surface fluxes. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the 3-month long output used by
\citet{rocha_etal2016}.

A key aspect of the LLC simulations is that they were forced by
the 16 most-significant tidal components.
Because barotropic tides interact with topography and generate internal
tides that project onto mesoscales to submesoscales
\citep[e.g., ][]{rocha_etal2016}, tidal forcing fundamentally distinguishes our analysis
from other modeling studies of seasonality \citep{sasaki_etal2014,qiu_etal2014}. Details of the LLC simulations
are provided in the supplemental material.

@@

The bulk of vertical vorticity and strain rate are associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields suppresses super-inertial
motions and reduces the RMS
vorticity by 40$\%$ and the RMS strain by 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km dramatically reduces the RMS
vorticity and strain rate. The reduction in variance is about 80$\%$ in winter,
yielding RMS vorticity and strain  rate roughly consistent with the diagnostics from
AVISO gridded geostrophic velocities (compare red lines to black lines in figures \ref{fig2}a-b).
The picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late winter, drawing from the available
potential energy stored in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

@@

\section{Seasonality in governing submesoscale dynamics}
The results of figure \ref{fig2}a-c show that submesoscale surface
 variability stems from different dynamics in summer than in winter.
 To characterize these differences, we calculate
joint probability distributions (jPDF) of
vorticity-strain and vorticity-Laplacian of sea-surface height (Figure \ref{fig3}).

@@

Figure \ref{fig4}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the April and October spectra based on hourly (the solid curves)
velocity snapshots are nearly indistinguishable from each other within 95$\%$ confidence level.
Consistent with the results of
\citet{rocha_etal2016}, who analyzed a 5-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig4}a). But this suppression is dramatic in October, when the inertia-gravity waves peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

@@

\cite{dasaro1978} showed that the velocity of linear internal waves
in the mixed-layer strongly depends on the density jump at the mixed-layer
base, with largest mixed layer velocities when the jump is strongest.
In summer, the shallow mixed layer overlays a strong seasonal pycnocline,
and thus the internal waves projection onto the mixed-layer may be stronger
according to \cite{dasaro1978}'s arguments. An
alternative explanation is that the shape of the baroclinic modes
changes seasonally. In particular the baroclinic modes are
significantly more surface intensified in late summer/early fall
 (supplemental material).
If the internal wave source has weak seasonal dependence, then a strong
seasonal cycle in the near-surface expression of internal waves is expected.
Internal tide generation does not show a seasonal modulation \citep[e.g.,][]{alford2003},
and therefore one expects a strong seasonality in the near-surface
expression of internal tides and
other submesoscale inertia-gravity waves generated through internal tide interactions.
Near-inertial wave generation peaks in winter \citep[e.g.,][]{alford2003}, but
those waves project on large horizontal scales \citep[e.g, ][]{qi_etal1995}.

@@

\bibliographystyle{agufull08}
\bibliography{rocha_etal}

@@

%% ------------------------------------------------------------------------ %%
%%  REFERENCE LIST AND TEXT CITATIONS
%
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% Or,
%
% If you use BiBTeX for your references, please use the agufull08.bst file (available at % ftp://ftp.agu.org/journals/latex/journals/Manuscript-Preparation/) to produce your .bbl
% file and copy the contents into your paper here.
%
% Follow these steps:
% 1. Run LaTeX on your LaTeX file.
%
% 2. Make sure the bibliography style appears as \bibliographystyle{agufull08}. Run BiBTeX on your LaTeX
% file.
%
% 3. Open the new .bbl file containing the reference list and
%   copy all the contents into your LaTeX file here.
%
% 4. Comment out the old \bibliographystyle and \bibliography commands.
%
% 5. Run LaTeX on your new file before submitting.
%
% AGU does not want a .bib or a .bbl file. Please copy in the contents of your .bbl file here.

>>>>

\begin{abstract}
    Two new high-resolution numerical simulations with embedded tides show a
    strong modulation of near-surface dynamics at submesoscales
    (roughly 10-100 km) in the Kuroshio Extension. Consistent with recent studies, deep late-winter mixed layers
    are prone to baroclinic instabilities, and submesoscale turbulence
    prevails in late winter/early spring. While summertime
    re-stratification weakens submesoscale turbulence, it also enhances submesoscale inertia-gravity
     waves near the surface. In the Kuroshio Extension,
    inertia-gravity waves strongly dominate the submesoscale surface kinetic energy and
    sea-surface height variance in late summer/early fall.
\end{abstract}

@@

Recent interest in upper-ocean dynamics has focused on the strong seasonal
cycle of shallow baroclinic instabilities and their role in submesoscale (roughly 1-100 km)
turbulence and mesoscale (roughly 100-300 km) modulation \citep{sasaki_etal2014,qiu_etal2014,
callies_etal2015, thompson_etal2016,buckingham_etal2016}. Contemporary studies
have also shown that inertia-gravity waves contribute significantly
to the near-surface variability at submesoscales \citep{richman_etal2012,
buhler_etal2014,rocha_etal2016}, but their seasonality has not been investigated.

@@

Inertia-gravity waves, on the other hand, peak in late summer/early fall,
when the upper ocean is strongly stratified. Thus there exists a
strong seasonal modulation of upper-ocean submesoscale dynamics:
submesoscale turbulence dominates the upper-ocean dynamics in late winter/early,
whereas submesoscale inertia-gravity waves prevail in late summer/early fall.
Because submesoscale turbulence is weakest in late summer/early fall,
the present results indicate that inertia-gravity waves account for most of the
summertime submesoscale SSH variability.

\section{The LLC numerical simulations}
We use the output of two latitude-longitude polar cap (LLC)
realistic numerical simulations. The outputs
analyzed here, LLC2160 (1/24$^\circ$) and LLC4320 (1/48$^\circ$),   are
forward Massachusetts Institute of Technology general circulation model (MITgcm)
numerical solutions on a LLC grid  \citep{forget_etal2015} with
90-vertical-levels. The coarser-resolution LLC simulation was
spun up from an ECCO2 adjoint-method state estimate, constrained to millions
of observations from 2009 through 2011. Both simulations were forced by
tides and 6-hourly surface fluxes. The LLC2160
output spans two years from March 2011 to April 2013; the LLC4320 was spun up from
the LLC2160 simulation, spanning one year from September 2011 to September 2012.
The LLC4320 simulation is an extension of the 3-month long output used by
\citet{rocha_etal2016}.  Details of the LLC simulations are provided in the supplemental material.

A key aspect of the LLC2160 and LLC4320 simulations is that they were forced by
the 16 most-significant tidal components.
Because barotropic tides interact with topography and generate internal
tides that project onto mesoscales to submesoscales
\citep[e.g., ][]{rocha_etal2016}, tidal forcing fundamentally distinguishes our analysis
from other modeling studies of seasonality \citep{sasaki_etal2014,qiu_etal2014}.

@@

The bulk of vertical vorticity and strain rate are associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields suppresses super-inertial
motions and reduces the RMS
vorticity by 40$\%$ and the RMS strain by 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Indeed, most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km dramatically reduces the RMS
vorticity and strain rate. The reduction in variance is about 80$\%$ in winter,
yielding RMS vorticity and strain  rate roughly consistent with the diagnostics from
AVISO gridded geostrophic velocities (compare red lines to black lines in figures \ref{fig2}a-b).
The picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late winter, drawing from the available
potential energy stored in large lateral buoyancy gradients in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

@@

\section{Seasonality of submesoscale dynamics}
The results of figure \ref{fig2}a-c show that submesoscale surface
 variability stems from different dynamics in summer than in winter.
 To characterize these differences, we calculate
joint probability distributions (jPDF) of
vorticity-strain and vorticity-Laplacian of sea-surface height (Figure \ref{fig3}).

@@

Figure \ref{fig4}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the April and October spectra based on hourly (the solid curves)
velocity snapshots are nearly indistinguishable from each other within 95$\%$ confidence level.
Consistent with the results of
\citet{rocha_etal2016}, who analyzed a 3-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig4}a). But this suppression is dramatic in October, when the inertia-gravity waves peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

@@

\cite{dasaro1978} showed that the velocity of linear internal waves
in the mixed layer strongly depends on the density jump at the mixed-layer
base, with largest mixed layer velocities when the jump is strongest.
In summer, the shallow mixed layer overlays a strong seasonal pycnocline,
and thus the internal waves projection onto the mixed-layer may be stronger
according to \cite{dasaro1978}'s arguments. An
alternative explanation is that the shape of the baroclinic modes
changes seasonally. In particular the baroclinic modes are
significantly more surface intensified in late summer/early fall
 (supplemental material).
If the internal wave source has weak seasonal dependence, then a strong
seasonal cycle in the near-surface expression of internal waves is expected.
Internal tide generation does not show a seasonal modulation \citep[e.g.,][]{alford2003},
and therefore one expects a strong seasonality in the near-surface
expression of internal tides and
other submesoscale inertia-gravity waves generated through internal tide interactions.
Near-inertial wave generation peaks in winter \citep[e.g.,][]{alford2003}, but
those waves project on large horizontal scales \citep[e.g, ][]{qi_etal1995}.

@@

\bibliographystyle{agufull08}
%\bibliography{rocha_etal}

@@

%% ------------------------------------------------------------------------ %%
%%  REFERENCE LIST AND TEXT CITATIONS
%
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
\begin{thebibliography}{16}
\providecommand{\natexlab}[1]{#1}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi

\bibitem[{\textit{Alford}(2003)}]{alford2003}
Alford, M.~H. (2003), {Redistribution of energy available for ocean mixing by
  long-range propagation of internal wave}, \textit{Nature},
  \textit{423}(6936), 159--162.

\bibitem[{\textit{Buckingham et~al.}(2016)\textit{Buckingham, Naveira~Garabato,
  Thompson, Brannigan, Lazar, Marshall, George~Nurser, Damerell, Heywood, and
  Belcher}}]{buckingham_etal2016}
Buckingham, C.~E., A.~C. Naveira~Garabato, A.~F. Thompson, L.~Brannigan,
  A.~Lazar, D.~P. Marshall, A.~George~Nurser, G.~Damerell, K.~J. Heywood, and
  S.~E. Belcher (2016), {Seasonality of submesoscale flows in the ocean surface
  boundary layer}, \textit{Geophysical Research Letters}, \textit{43}(5),
  2118--2126.

\bibitem[{\textit{B{\"u}hler et~al.}(2014)\textit{B{\"u}hler, Callies, and
  Ferrari}}]{buhler_etal2014}
B{\"u}hler, O., J.~Callies, and R.~Ferrari (2014), {Wave--vortex decomposition
  of one-dimensional ship-track data}, \textit{Journal of Fluid Mechanics},
  \textit{756}, 1007--1026.

\bibitem[{\textit{Callies et~al.}(2015)\textit{Callies, Ferrari, Klymak, and
  Gula}}]{callies_etal2015}
Callies, J., R.~Ferrari, J.~M. Klymak, and J.~Gula (2015), {Seasonality in
  submesoscale turbulence}, \textit{Nature communications}, \textit{6}.

\bibitem[{\textit{Callies et~al.}(2016)\textit{Callies, Flierl, Ferrari, and
  Fox-Kemper}}]{callies_etal2016}
Callies, J., G.~Flierl, R.~Ferrari, and B.~Fox-Kemper (2016), {The role of
  mixed-layer instabilities in submesoscale turbulence}, \textit{Journal of
  Fluid Mechanics}, \textit{788}, 5--41.

\bibitem[{\textit{Capet et~al.}(2008)\textit{Capet, McWilliams, Molemaker, and
  Shchepetkin}}]{capet_etal2008a}
Capet, X., J.~C. McWilliams, M.~J. Molemaker, and A.~Shchepetkin (2008),
  {Mesoscale to submesoscale transition in the California Current System. Part
  I: Flow structure, eddy flux, and observational tests}, \textit{Journal of
  Physical Oceanography}, \textit{38}(1), 29--43.

\bibitem[{\textit{D'Asaro}(1978)}]{dasaro1978}
D'Asaro, E.~A. (1978), {Mixed layer velocities induced by internal wave},
  \textit{Journal of Geophysical Research: Oceans}, \textit{83}(C5),
  2437--2438.

\bibitem[{\textit{Forget et~al.}(2015)\textit{Forget, Campin, Heimbach, Hill,
  Ponte, and Wunsch}}]{forget_etal2015}
Forget, G., J.-M. Campin, P.~Heimbach, C.~Hill, R.~Ponte, and C.~Wunsch (2015),
  {ECCO version 4: an integrated framework for non-linear inverse modeling and
  global ocean state estimation}, \textit{Geosci. Model Dev.}, \textit{8},
  3071--3104.

\bibitem[{\textit{McWilliams}(2016)}]{mcwilliams2016}
McWilliams, J.~C. (2016), {Submesoscale currents in the ocean}, in
  \textit{Proc. R. Soc. A}, vol. 472, p. 20160117, The Royal Society.

\bibitem[{\textit{Qi et~al.}(1995)\textit{Qi, De~Szoeke, Paulson, and
  Eriksen}}]{qi_etal1995}
Qi, H., R.~A. De~Szoeke, C.~A. Paulson, and C.~C. Eriksen (1995), {The
  structure of near-inertial waves during ocean storms}, \textit{Journal of
  Physical Oceanography}, \textit{25}, 2853--2871.

\bibitem[{\textit{Qiu et~al.}(2014)\textit{Qiu, Chen, Klein, Sasaki, and
  Sasai}}]{qiu_etal2014}
Qiu, B., S.~Chen, P.~Klein, H.~Sasaki, and Y.~Sasai (2014), {Seasonal mesoscale
  and submesoscale eddy variability along the North Pacific Subtropical
  Countercurrent}, \textit{Journal of Physical Oceanography}, \textit{44}(12),
  3079--3098.

\bibitem[{\textit{Richman et~al.}(2012)\textit{Richman, Arbic, Shriver,
  Metzger, and Wallcraft}}]{richman_etal2012}
Richman, J.~G., B.~K. Arbic, J.~F. Shriver, E.~J. Metzger, and A.~J. Wallcraft
  (2012), {Inferring dynamics from the wavenumber spectra of an eddying global
  ocean model with embedded tides}, \textit{Journal of Geophysical Research:
  Oceans}, \textit{117}(C12).

\bibitem[{\textit{Rocha et~al.}(2016)\textit{Rocha, Chereskin, Gille, and
  Menemenlis}}]{rocha_etal2016}
Rocha, C.~B., T.~K. Chereskin, S.~T. Gille, and D.~Menemenlis (2016),
  {Mesoscale to submesoscale wavenumber spectra in Drake Passage},
  \textit{Journal of Physical Oceanography}, \textit{46}, 601--620.

\bibitem[{\textit{Sasaki et~al.}(2014)\textit{Sasaki, Klein, Qiu, and
  Sasai}}]{sasaki_etal2014}
Sasaki, H., P.~Klein, B.~Qiu, and Y.~Sasai (2014), {Impact of oceanic-scale
  interactions on the seasonal modulation of ocean dynamics by the atmosphere},
  \textit{Nature communications}, \textit{5}.

\bibitem[{\textit{Shcherbina et~al.}(2013)\textit{Shcherbina, D'Asaro, Lee,
  Klymak, Molemaker, and McWilliams}}]{shcherbina_etal2013}
Shcherbina, A.~Y., E.~A. D'Asaro, C.~M. Lee, J.~M. Klymak, M.~J. Molemaker, and
  J.~C. McWilliams (2013), {Statistics of vertical vorticity, divergence, and
  strain in a developed submesoscale turbulence field}, \textit{Geophysical
  Research Letters}, \textit{40}(17), 4706--4711.

\bibitem[{\textit{Thompson et~al.}(2016)\textit{Thompson, Lazar, Buckingham,
  Naveira~Garabato, Damerell, and Heywood}}]{thompson_etal2016}
Thompson, A.~F., A.~Lazar, C.~Buckingham, A.~C. Naveira~Garabato, G.~M.
  Damerell, and K.~J. Heywood (2016), {Open-ocean submesoscale motions: A full
  seasonal cycle of mixed layer instabilities from gliders}, \textit{Journal of
  Physical Oceanography}, \textit{46}(4), 1285--1307.

\end{thebibliography}

% Or,
%
% If you use BiBTeX for your references, please use the agufull08.bst file (available at % ftp://ftp.agu.org/journals/latex/journals/Manuscript-Preparation/) to produce your .bbl
% file and copy the contents into your paper here.
%
% Follow these steps:
% 1. Run LaTeX on your LaTeX file.
%
% 2. Make sure the bibliography style appears as \bibliographystyle{agufull08}. Run BiBTeX on your LaTeX
% file.
%
% 3. Open the new .bbl file containing the reference list and
%   copy all the contents into your LaTeX file here.
%
% 4. Comment out the old \bibliographystyle and \bibliography commands.
%
% 5. Run LaTeX on your new file before submitting.
%
% AGU does not want a .bib or a .bbl file. Please copy in the contents of your .bbl file here.
